---
title: "Sampling_w_replacement_Pollock_lngs"
output: html_document
date: "2026-03-17"
editor_options: 
  chunk_output_type: console
---

Based off of conversations with Matt and Curry: 

We have a subset of the total catch of age-0 pollock that are measured for SL each station/year in the NBS/SEBS
We want to assign lengths to the rest of the catch ("catch-weighting") so that we can account for the 
proportion of fish that we sampled (i.e., catch numbers) and get a true distribution of lengths of fish that were 
sampled and counted at each station. 

A way to do this is to use "sample" in base R
take the known lengths (length) and the number of unknown lengths (total catch number - number of sampled lengths)
sample(x = lengths, size = catch_number - measured_number, replace = TRUE)

However, this is assigning lengths based on the sample proportion and those simulated numbers will change with each re-run of "sample"
A way to deal with this is to bootstrap this sample procedure (by station/year) and obtain a sense of the 
variability of the data. If all we cared about was the mean length we could average the bootstrapped length assignment, 
but we want to use the assigned lengths as input for a spatiotemporal model like sdmTMB. 

Ideally - we would use these bootstrapped data and EACH TIME we bootstrapped the data we would REFIT the sdmTMB model 
to see how the variable the output of that model is (e.g., variability in terms of parameter estimates etc). However, this in particular, feels 
infesiable to do for every station every year. 

This sample procedure will likely be pretty consistent in assignment across bootstraps if the number of sampled/measured fish is large in proportion to the number of unsampled/measured fish (e.g., if we measured 50 fish and there are only 25 unmeasured fish). However, it may be more variable if the ratio of unsampled/measured fish is high (e.g., if we measured 50 fish and there are 1000 unmeasure fish). 

Therefore, instead of bootstrapping all stations all years we could look at a few instances of mean/sd lengths depending on the proportion of unsampled/measured fishes. 

```{r}
library(tidyverse)
library(purrr)
```


Downstream questions I have after we've assigned the length to the catch numbers:
- what to do about instances where we have no length subsamples, but we DO have catch numbers for our sdmTMB model.

Points to consider - pollock were clearly there (becasue of the positive catch), but there's no length to base the extrapolation on? 
How do we deal with this in the model? Pretend like the station wasn't sampled? 

- Right now the catch weight data I'm reading in below does NOT include all the catch data but only includes catch where we have subsampled lengths for

```{r}
catch <- read.csv("Data/a0pollock_catch_w_length_samples.csv")
lngs <- read.csv("Data/a0pollock_corrected_lngs.csv")
```


Steps - 
1. look at how many length samples are at each sampling event (station) and what proportion of the catch those represent
2. set up a bootstrap of the sample() with replacement function that calculates mean and sd of the TOTAL sample related to the proportion of unsampled catch
Therefore looking at a few instances of mean/sd lengths depending on the proportion of unsampled/measured fishes. 

# 1. Proportion of sample/unsampled

```{r}
nrow(lngs)
nrow(catch)

# we have a couple db issues where some lengths were included or not included well fix them here until the db fix gets pushed
cor.lng <- 0.29353 + 0.92359*93 # length of the fish in the origiinal records
lngs <- lngs %>%
  mutate(corrected_length = ifelse(is.na(corrected_length) & station_id == "20060201015",
                                   cor.lng, corrected_length)) %>%
  filter(!(is.na(corrected_length & station_id == "20100501004")))

# get sample sizes from lng samples
lng.sample.size <- lngs %>%
  group_by(station_id) %>%
  dplyr::summarise(num_lng = n()) # 1139

catch_prop <-left_join(catch, lng.sample.size) %>%
  mutate(total_catch_number = ifelse(station_id == "20110304024" & total_catch_number == 39, 40, total_catch_number)) %>%
  group_by(station_id) %>%
  mutate(proportion_sampled = num_lng / total_catch_number) %>%
  mutate(station_id = as.character(station_id))

# visualize it
range(catch_prop$proportion_sampled) # okay good we dont have more than 1 

more.lng.than.catch <- filter(catch_prop, proportion_sampled > 1) # okay this is an instance of 40 lng in the specimen records and 39 in the catch 
# to be honest I will change this (above) to make it match our lng records cause those were the touched fish
# this likely will mess up weights but we're not using that so its okay
filter(catch_prop, station_id == "20110304024") # dyson survey

hist(catch_prop$proportion_sampled) # okay looking at this there's mainly low proportion of catch sampled (and a few instances of 1:1 where every fish was measured
# and we won't be bootstrapping over)

# now we have fewer instances (780 obs) where we need to bootstrap over yay! 
catch_prop_sub <- filter(catch_prop, proportion_sampled != 1)
table(catch_prop_sub$sample_year)
# okay here we have the number of stations by years that we'll be resample/extrapolating over
```

2. set up a bootstrap of the sample() with replacement function that calculates mean and sd of the TOTAL sample related to the proportion of unsampled catch
Therefore looking at a few instances of mean/sd lengths depending on the proportion of unsampled/measured fishes. 

# 2. Bootstrap sample()

sample(x = lengths, size = catch_number - measured_number, replace = TRUE)

Using map and lists is faster than rebuilding the df using rbind, this still takes awhile to run 

```{r}
set.seed(033026)

# Pre-split data to avoid repeated filtering
lngs_split <- split(lngs, lngs$station_id)
catch_split <- split(catch_prop_sub, catch_prop_sub$station_id)

dataset_list <- map(names(catch_split), function(i) {
  
  # Extract relevant data for this station
  lngs_i <- lngs_split[[i]]$corrected_length
  catch_i <- catch_split[[i]]
  
  n_to_sample <- catch_i$total_catch_number - catch_i$num_lng
  n_total <- catch_i$total_catch_number
  
  # Inner Loop: Use map_df to return a data frame directly
  boot_df <- map_df(1:400, function(b) {
    # Bootstrap sample (working with vectors, not data frames)
    new_lngs <- sample(lngs_i, size = n_to_sample, replace = TRUE)
    all_lngs <- c(new_lngs, lngs_i)
    
    # Calculate metrics
    m <- mean(all_lngs)
    s <- sd(all_lngs)
    
    # Return one row
    data.frame(station_id = i, mean = m, sd = s, 
               num = n_total, se = s/sqrt(n_total), 
               bootstrap_it = b, sample_year = catch_i$sample_year)
  })
  
  return(boot_df)
})

# Final assembly
final_dataset <- bind_rows(dataset_list)
```

Now lets compare how the variable the means were

so basically we want to know what the variability of the means by station adn proportion sampled
```{r}
tmp <- final_dataset %>%
  group_by(station_id) %>%
  dplyr::summarise(mean_sd = sd(mean),
                   ) %>%
  left_join(catch_prop)

scale_means <- final_dataset %>%
  group_by(station_id) %>%
  dplyr::summarise(scaled_mean = scale(mean)) %>%
  left_join(catch_prop)


ggplot() +
  geom_point(aes(x = proportion_sampled, y = mean_sd), data = tmp) +
  labs(x = "Proportion of total catch subsampled and lengthed", y = "Standard deviation of bootstrapped means") +
  theme_bw()

range(tmp$mean_sd, na.rm = T)
range(tmp$mean_sd)
# why are there standard deviations that have NAs?
test <-filter(tmp, is.na(mean_sd))

# all of the means at these two stations were nas
na.means<-subset(final_dataset, station_id %in% test$station_id)

lng.nas <- subset(lngs, station_id %in% test$station_id) %>%
  filter(is.na(corrected_length))

catch.nas <- subset(catch_prop_sub, station_id %in% test$station_id)

# okay we've identified the issues, we only have 55 measurements in 2010 which means
# that someone just counted wrong and added in too many rows in specimen table 
lng.2010 <- subset(lngs, cruise_id == "EE1001" & station_id == "20100501004")
lng.2006 <- subset(lngs, station_id == "20060201015")

```

a couple of the bootstrapped points have mean_sd that are quite high ie above 1.25
```{r}
ggplot() +
  geom_point(aes(x = proportion_sampled, y = mean_sd,
                 fill = mean_sd > 1.25), data = tmp, pch = 21) +
  labs(x = "Proportion of total catch subsampled and lengthed", y = "Standard deviation of bootstrapped means") +
  theme_bw()

high.variability <- filter(tmp, mean_sd > 1.25) # 7 stations with super variable standard deviatiosn of mean estimates


range(tmp$proportion_sampled)
range(high.variability$proportion_sampled) # surprisingly its not like the proportion sampled is all that low

variable.lngs <- subset(lngs, station_id %in% c(high.variability$station_id))

variable.lngs %>%
  mutate(station_id = as.character(station_id)) %>%
  group_by(sample_year, station_id) %>%
  dplyr::summarize(low.range = min(corrected_length),
                   high.range = max(corrected_length),
                   num = n()) %>%
  left_join(high.variability) %>%
  dplyr::select(sample_year:mean_sd, lhs_code, common_name, total_catch_number, total_catch_weight, catch_notes, num_lng, proportion_sampled)

# what if its not really the proportion sampled that causes the issues but the absolute number of lengths taken?

ggplot() +
  geom_point(aes(x = num_lng, y = mean_sd,
                 fill = mean_sd > 1.25), data = tmp, pch = 21) +
  labs(x = "Absolute number of lengths sampled", y = "Standard deviation of bootstrapped means") +
  theme_bw()

# this looks like the case. 
# if we remove the points for the sake of scale does this still look okay
tmp %>%
  filter(mean_sd < 1.25) %>%
  ggplot() +
  geom_point(aes(x = num_lng, y = mean_sd)) +
  labs(x = "Absolute number of lengths sampled", y = "Standard deviation of bootstrapped means") +
  theme_bw() +
  geom_smooth(aes(x = num_lng, y = mean_sd))


tmp %>%
  filter(mean_sd < 1.25) %>%
  ggplot() +
  geom_point(aes(x = proportion_sampled, y = mean_sd)) +
  labs(x = "Proportion of total catch subsampled and lengthed", y = "Standard deviation of bootstrapped means") +
  geom_smooth(aes(x = proportion_sampled, y = mean_sd)) +
  theme_bw()
```

what if we looked at scaled mean values
```{r}
scale_means %>%
  ggplot() +
  geom_point(aes(x = proportion_sampled, y = scaled_mean[,1]))

# honestly looks pretty decent if we sample more than 10% of the catch

```



