---
title: "Sampling_w_replacement_Pollock_lngs"
output: html_document
date: "2026-03-17"
editor_options: 
  chunk_output_type: console
---

Based off of conversations with Matt and Curry: 

We have a subset of the total catch of age-0 pollock that are measured for SL each station/year in the NBS/SEBS
We want to assign lengths to the rest of the catch ("catch-weighting") so that we can account for the 
proportion of fish that we sampled (i.e., catch numbers) and get a true distribution of lengths of fish that were 
sampled and counted at each station. 

A way to do this is to use "sample" in base R
take the known lengths (length) and the number of unknown lengths (total catch number - number of sampled lengths)
sample(x = lengths, size = catch_number - measured_number, replace = TRUE)

However, this is assigning lengths based on the sample proportion and those simulated numbers will change with each re-run of "sample"
A way to deal with this is to bootstrap this sample procedure (by station/year) and obtain a sense of the 
variability of the data. If all we cared about was the mean length we could average the bootstrapped length assignment, 
but we want to use the assigned lengths as input for a spatiotemporal model like sdmTMB. 

Ideally - we would use these bootstrapped data and EACH TIME we bootstrapped the data we would REFIT the sdmTMB model 
to see how the variable the output of that model is (e.g., variability in terms of parameter estimates etc). However, this in particular, feels 
infesiable to do for every station every year. 

This sample procedure will likely be pretty consistent in assignment across bootstraps if the number of sampled/measured fish is large in proportion to the number of unsampled/measured fish (e.g., if we measured 50 fish and there are only 25 unmeasured fish). However, it may be more variable if the ratio of unsampled/measured fish is high (e.g., if we measured 50 fish and there are 1000 unmeasure fish). 

Therefore, instead of bootstrapping all stations all years we could look at a few instances of mean/sd lengths depending on the proportion of unsampled/measured fishes. 

# library
```{r}
library(tidyverse)
library(purrr)
library(sdmTMB)
library(sf)
library(concaveman)
library(visreg)

#getNOAABathy lines
library(marmap)
#for plotting maps and projections
library(maps)
library(mapdata)
library(mapproj)
```


Downstream questions I have after we've assigned the length to the catch numbers:
- what to do about instances where we have no length subsamples, but we DO have catch numbers for our sdmTMB model.

Points to consider - pollock were clearly there (becasue of the positive catch), but there's no length to base the extrapolation on? 
How do we deal with this in the model? Pretend like the station wasn't sampled? 

- Right now the catch weight data I'm reading in below does NOT include all the catch data but only includes catch where we have subsampled lengths for

```{r}
catch <- read.csv("Data/a0pollock_catch_w_length_samples_041326.csv")
lngs <- read.csv("Data/a0pollock_corrected_lngs_041326.csv")

# direct.lng <- join_event_fish(start_year = 2003, end_year = 2025, survey_region = c("NBS", "SEBS"),
#                               tsn = 934083, gear = "CAN", trawl_method = "S")
# 
# direct.lng.a0 <- filter(direct.lng, lhs_code == "A0")
# # interestingly 
# table(is.na(direct.lng.a0$length))
```

# spatial and bathy files
```{r}
# #read in Alaska polygon for mapping
# US <- map_data("world2Hires",region="USA")
# AK <- subset(US,subregion == "Alaska")
# USSR <- map_data("world2Hires",region="USSR")
# BS <- rbind(AK,USSR)
# #transforms longitudes (you could also transform the longs in the data file)
# BS$long <- BS$long-360
# 
# BS.crop <- filter(BS, long < -155, long > - 175,
#                       lat < 65)
# 
# BS2 <- st_as_sf(BS, coords = c("long", "lat"), crs = 4326) %>%
#   st_polygon

# get maps data
# bs_map <- maps::map("world2Hires", regions = c("USA:Alaska", "USSR"), 
#                     fill = TRUE, plot = FALSE) %>%
#   st_as_sf(crs = 26902) # i think its 4326, but also trying out 26902

sf_use_s2(TRUE)

# need to use world not world2hires cause that one is hella old and doesn't plot correctly
bs_map <- maps::map("world", regions = c("USA:Alaska", "Russia"), 
                    fill = TRUE, plot = FALSE, wrap = c(0, 360)) %>%
  st_as_sf(crs = 4326) 

# Use bounding box, have to add 360 to the bbox coords
# xmin = 185 (-175), xmax = 202 (-155)
bbox_360 <- st_bbox(c(xmin = 180, xmax = 210, ymin = 50, ymax = 68), 
                    crs = st_crs(bs_map))

# crop 
# Using st_make_valid helps w/ messy polys
bs_cropped <- bs_map %>% 
  st_make_valid() %>% 
  st_intersection(st_as_sfc(bbox_360)) %>%
  st_transform(crs = 32602) # Move to UTM Zone 2N

#read in bathymetry file.  
bathline <- getNOAA.bathy(-180,-155,54,68)

bathline.xyz <- as.xyz(bathline)

# plot
# ggplot() +
#   geom_sf(data = bs_cropped, fill = "antiquewhite", color = "black") +
#   theme_minimal() +
#   geom_contour(data = bathline, mapping = aes(x = x + 360, y = y, z = z),
#                breaks = c(-20, -50, -100))
```


Steps - 
1. look at how many length samples are at each sampling event (station) and what proportion of the catch those represent
2. set up a bootstrap of the sample() with replacement function that calculates mean and sd of the TOTAL sample related to the proportion of unsampled catch
Therefore looking at a few instances of mean/sd lengths depending on the proportion of unsampled/measured fishes. 

# 1. Proportion of sample/unsampled

```{r}
nrow(lngs)
nrow(catch)

table(is.na(lngs$corrected_length))
na.lngs <- filter(lngs, is.na(corrected_length))
# In 2011 the specimen record said that there was an illogical length so it was deleted but the record kept
#' 2012 it looks like possibily a stub record although not confirmed of a specimen that was enetered but not measured
#' (was an arctic cod but go re-assigned post processing with a voucher sample)
#' 2018 CLAMS broke and caused a lot of angst andrew tried to reconstruct it then but didn't work
#' The NAs represent the number of samples that were collected, but no lengths and that the best we can do
# write.csv(na.lngs, "Data/na_lengths_a0_pollock.csv")

# SEPERATE FROM THE NAs above: 
# we have a couple db issues where some lengths were included or not included well fix them here until the db fix gets pushed
cor.lng <- 0.29353 + 0.92359*93 # length of the fish in the origiinal records
lngs <- lngs %>%
  mutate(corrected_length = ifelse(is.na(corrected_length) & station_id == "20060201015",
                                   cor.lng, corrected_length)) %>%
  filter(!(is.na(corrected_length & station_id == "20100501004"))) %>%
  # remove any remaining NAs
  filter(!(is.na(corrected_length)))




# get sample sizes from lng samples
lng.sample.size <- lngs %>%
  group_by(station_id) %>%
  dplyr::summarise(num_lng = n()) # 1139

catch_prop <-left_join(catch, lng.sample.size) %>%
  mutate(total_catch_number = ifelse(station_id == "20110304024" & total_catch_number == 39, 40, total_catch_number)) %>%
  group_by(station_id) %>%
  mutate(proportion_sampled = num_lng / total_catch_number) %>%
  mutate(station_id = as.character(station_id))

# visualize it
range(catch_prop$proportion_sampled) # okay good we dont have more than 1 

more.lng.than.catch <- filter(catch_prop, proportion_sampled > 1) # okay this is an instance of 40 lng in the specimen records and 39 in the catch 
# to be honest I will change this (above) to make it match our lng records cause those were the touched fish
# this likely will mess up weights but we're not using that so its okay
filter(catch_prop, station_id == "20110304024") # dyson survey

hist(catch_prop$proportion_sampled) # okay looking at this there's mainly low proportion of catch sampled (and a few instances of 1:1 where every fish was measured
# and we won't be bootstrapping over)

# now we have fewer instances (780 obs) where we need to bootstrap over yay! 
catch_prop_sub <- filter(catch_prop, proportion_sampled != 1)
table(catch_prop_sub$sample_year)
# okay here we have the number of stations by years that we'll be resample/extrapolating over
```

2. set up a bootstrap of the sample() with replacement function that calculates mean and sd of the TOTAL sample related to the proportion of unsampled catch
Therefore looking at a few instances of mean/sd lengths depending on the proportion of unsampled/measured fishes. 

# 2. Bootstrap sample()

sample(x = lengths, size = catch_number - measured_number, replace = TRUE)

Using map and lists is faster than rebuilding the df using rbind, this still takes awhile to run 

```{r}
# Pre-split data to avoid repeated filtering
lngs_split <- split(lngs, lngs$station_id)
catch_split <- split(catch_prop_sub, catch_prop_sub$station_id)
```

```{r eval=FALSE, include=FALSE}
set.seed(033026)

dataset_list <- map(names(catch_split), function(i) {
  
  # Extract relevant data for this station
  lngs_i <- lngs_split[[i]]$corrected_length
  catch_i <- catch_split[[i]]
  
  n_to_sample <- catch_i$total_catch_number - catch_i$num_lng
  n_total <- catch_i$total_catch_number
  
  # Inner Loop: Use map_df to return a data frame directly
  boot_df <- map_df(1:400, function(b) {
    # Bootstrap sample (working with vectors, not data frames)
    new_lngs <- sample(lngs_i, size = n_to_sample, replace = TRUE)
    all_lngs <- c(new_lngs, lngs_i)
    
    # Calculate metrics
    m <- mean(all_lngs)
    s <- sd(all_lngs)
    
    # Return one row
    data.frame(station_id = i, mean = m, sd = s, 
               num = n_total, se = s/sqrt(n_total), 
               bootstrap_it = b, sample_year = catch_i$sample_year)
  })
  
  return(boot_df)
})

# Final assembly
final_dataset <- bind_rows(dataset_list)
```

Now lets compare how the variable the means were

so basically we want to know what the variability of the means by station adn proportion sampled
```{r eval=FALSE, include=FALSE}
tmp <- final_dataset %>%
  group_by(station_id) %>%
  dplyr::summarise(mean_sd = sd(mean),
                   ) %>%
  left_join(catch_prop)

scale_means <- final_dataset %>%
  group_by(station_id) %>%
  dplyr::summarise(scaled_mean = scale(mean)) %>%
  left_join(catch_prop)


ggplot() +
  geom_point(aes(x = proportion_sampled, y = mean_sd), data = tmp) +
  labs(x = "Proportion of total catch subsampled and lengthed", y = "Standard deviation of bootstrapped means") +
  theme_bw()

range(tmp$mean_sd, na.rm = T)
range(tmp$mean_sd)
# why are there standard deviations that have NAs?
test <-filter(tmp, is.na(mean_sd))

# all of the means at these two stations were nas
na.means<-subset(final_dataset, station_id %in% test$station_id)

lng.nas <- subset(lngs, station_id %in% test$station_id) %>%
  filter(is.na(corrected_length))

catch.nas <- subset(catch_prop_sub, station_id %in% test$station_id)

# okay we've identified the issues, we only have 55 measurements in 2010 which means
# that someone just counted wrong and added in too many rows in specimen table 
lng.2010 <- subset(lngs, cruise_id == "EE1001" & station_id == "20100501004")
lng.2006 <- subset(lngs, station_id == "20060201015")

```

a couple of the bootstrapped points have mean_sd that are quite high ie above 1.25
```{r eval=FALSE, include=FALSE}
ggplot() +
  geom_point(aes(x = proportion_sampled, y = mean_sd,
                 fill = mean_sd > 1.25), data = tmp, pch = 21) +
  labs(x = "Proportion of total catch subsampled and lengthed", y = "Standard deviation of bootstrapped means") +
  theme_bw()

high.variability <- filter(tmp, mean_sd > 1.25) # 7 stations with super variable standard deviatiosn of mean estimates


range(tmp$proportion_sampled)
range(high.variability$proportion_sampled) # surprisingly its not like the proportion sampled is all that low

variable.lngs <- subset(lngs, station_id %in% c(high.variability$station_id))

variable.lngs %>%
  mutate(station_id = as.character(station_id)) %>%
  group_by(sample_year, station_id) %>%
  dplyr::summarize(low.range = min(corrected_length),
                   high.range = max(corrected_length),
                   num = n()) %>%
  left_join(high.variability) %>%
  dplyr::select(sample_year:mean_sd, lhs_code, common_name, total_catch_number, total_catch_weight, catch_notes, num_lng, proportion_sampled)

# what if its not really the proportion sampled that causes the issues but the absolute number of lengths taken?

ggplot() +
  geom_point(aes(x = num_lng, y = mean_sd,
                 fill = mean_sd > 1.25), data = tmp, pch = 21) +
  labs(x = "Absolute number of lengths sampled", y = "Standard deviation of bootstrapped means") +
  theme_bw()

# this looks like the case. 
# if we remove the points for the sake of scale does this still look okay
tmp %>%
  filter(mean_sd < 1.25) %>%
  ggplot() +
  geom_point(aes(x = num_lng, y = mean_sd)) +
  labs(x = "Absolute number of lengths sampled", y = "Standard deviation of bootstrapped means") +
  theme_bw() +
  geom_smooth(aes(x = num_lng, y = mean_sd))


tmp %>%
  filter(mean_sd < 1.25) %>%
  ggplot() +
  geom_point(aes(x = proportion_sampled, y = mean_sd)) +
  labs(x = "Proportion of total catch subsampled and lengthed", y = "Standard deviation of bootstrapped means") +
  geom_smooth(aes(x = proportion_sampled, y = mean_sd)) +
  theme_bw()


# what if we looked at scaled mean values
scale_means %>%
  ggplot() +
  geom_point(aes(x = proportion_sampled, y = scaled_mean[,1]))

# honestly looks pretty decent if we sample around/more than 10% of the catch

```

## 2.1 All lengths returned instead of summary statistics

```{r eval=FALSE, include=FALSE}
set.seed(033026)

# Pre-split data to avoid repeated filtering
lngs_split <- split(lngs, lngs$station_id)
catch_split <- split(catch_prop_sub, catch_prop_sub$station_id)

length_list <- map(names(catch_split), .progress = "Processing stations", function(i) {
  
  # Extract relevant data for this station
  lngs_i <- lngs_split[[i]]$corrected_length
  catch_i <- catch_split[[i]]
  
  n_to_sample <- catch_i$total_catch_number - catch_i$num_lng
  n_total <- catch_i$total_catch_number
  
  # Inner Loop: Use map_df to return a data frame directly
  boot_lengths <- map_df(1:20, function(b) {
    # Bootstrap sample (working with vectors, not data frames)
    new_lngs <- sample(lngs_i, size = n_to_sample, replace = TRUE)
    #all_lngs <- c(new_lngs, lngs_i)
    
    # # Calculate metrics
    # m <- mean(all_lngs)
    # s <- sd(all_lngs)
    
    # Return one row
    data.frame(station_id = i, 
               sample_year = catch_i$sample_year,
               bootstrap_it = b, 
               length = c(new_lngs, lngs_i)
               )
  })
  
  return(boot_lengths)
})

# Final assembly
length_dataset <- bind_rows(length_list)
```

how is this organized again?
```{r eval=FALSE, include=FALSE}
class(length_dataset)
names(length_dataset)

head(length_dataset)
str(length_dataset)

unique(length_list[[1]]$sample_year)# so each list of list is a specific station within a year that is 
unique(length_list[[1]]$bootstrap_it) # boot strapped 20 times. 
unique(length_list[[1]]$station_id)
nrow(length_list[[1]])

lngs_split$`20030101004`
catch_split$`20030101004`$num_lng # 10 actual lengths
catch_split$`20030101004`$total_catch_number # 20 in total sampled
catch_split$`20030101004`$proportion_sampled # so half measured
# i.e. for 20030101004 we should be estimating 10 additional lengths
# so 20 * 20 (cause bootstrapping 20x) means for this station its 400 rows long

length_list[1]
ggplot((length_list[[1]])) +
  geom_histogram(aes(x = length)) + 
  facet_wrap(~bootstrap_it)
```

## 2.2 doy to lengths

Get the doy data added in with the length data
```{r eval=FALSE, include=FALSE}

# add julian day to catch data
catch2 <- catch_prop_sub %>%
  mutate(date = as.Date(haul_date),
         doy = yday(date),
         station_id = as.character(station_id)) %>% 
  dplyr::select(sample_year, station_id, eq_latitude, eq_longitude, date, doy)

# whats the range in data, should we be including each doy or aggregate into weeks? 
# weeks may be a good option if the sdm is taking a long time to run
range(catch2$doy); max(catch2$doy) - min(catch2$doy) # survey data spans 54 days

# for now lets just take the first iteration of the bootstrap to play with a "full" dataset of lengths
length_it1 <- length_dataset %>%
  filter(bootstrap_it == 1)

# length_dat <- length_it1 %>%
#   left_join(catch2, by = c("station_id", "sample_year")) %>%
#   dplyr::select(-bootstrap_it)
```


# 3. sdmTMB w/ lengths

Create all the parts you'll need to fit and predict the sdmTMB model:
add utm to data, make mesh, create prediction grid

```{r eval=FALSE, include=FALSE}
# subset the length data to one iteration 
length_dat <- length_it1 %>%
  filter(sample_year %in% c(2018:2025)) %>%
  left_join(catch2, by = c("station_id", "sample_year")) %>%
  dplyr::select(-bootstrap_it)

# add utm to data
lng_utm <- length_dat %>%
  #dplyr::select(sample_year, station_id, eq_latitude, eq_longitude) %>%
  add_utm_columns(., ll_names = c("eq_longitude", "eq_latitude"),
                  ll_crs = 4326, utm_names = c("X", "Y"), 
                  units = "km") # uses crs 32602

head(lng_utm)

crs <- st_crs(lng_utm)
# make mesh
mesh <- make_mesh(lng_utm, c("X", "Y"), n_knots = 50)
plot(mesh)
# create prediction grid
## create sf object
lng_sf <- lng_utm %>%
  ungroup() %>%
  dplyr::select(c(X, Y, eq_latitude, eq_longitude)) %>%
  mutate(X = X*1000, Y = Y*1000) %>% # convert to meters for crs 32602
  st_as_sf(coords = c("X", "Y"), crs = 32602)

## create convex hull
convex <- concaveman(lng_sf, concavity = 3)

## add buffer to edges
convex_buff <- st_buffer(convex, dist = 15e3) # dist in meters (we think)
plot(convex_buff) # looks about right

## make grid
# make grid, created a cell size of 10,000 x 10,000 m (10 x 10 km) with poitns in center
# need to check units on this... 
grid <- st_make_grid(convex_buff, cellsize = 10e3, what = "centers") %>%
  st_intersection(convex_buff)

## change back to df
bs.grid <- do.call(rbind, st_geometry(grid)) %>%
  as_tibble() %>% setNames(c("X", "Y")) %>%
  # convert df m back to km
  mutate(X = X/1e3, Y = Y/1e3)
plot(bs.grid)

## replicate df across years and include any other variables we inlcude in our fit of the model
## i.e., depth tmeperature etc
bs.grid.year <- replicate_df(bs.grid, "sample_year", unique(lng_utm$sample_year))

bs.grid.year.doy <- replicate_df(bs.grid.year, "doy", unique(lng_utm$doy))
```


Now we can fit the sdmTMB model for lengths

What should we be considering - Lauren rogers used a gaussian distribution for similar length data from the larval samples
but its not a true normal distribution... 

following rogers et al., they investigated interannual changes in mean size by fitting models with year as factor and extracting conditional year effects for day of year (mean sampling date)

so i think what this means is that for each station we might need to include mean size? rather than all data? 
```{r eval=FALSE, include=FALSE}
# in order to 
fit1 <- sdmTMB(
  data = lng_utm,
  # model formula
  log(length) ~ 0 + s(doy) + as.factor(sample_year),
  # add mesh
  mesh = mesh, 
  # we want to fit a spatiotemporal model with iid so we need to say which is the time category
  time = "sample_year", 
  # family important.... start with gaussian, then lognormal 
  #family = gaussian(link = "identity"),
  family = lognormal(link = "log"), 
  spatial = "on",
  # each year independentally estimated
  spatiotemporal = "IID",
  # we want ansiotropy because we want to assume that the decomposition of Matern distance isn't
  # necessarily the same EW as NS. Esp cause in the BS we have the shelf patterns that can drive 
  # species distributions
  anisotropy = TRUE
)
```


## model checks
```{r eval=FALSE, include=FALSE}
# check the standard checks that the model has
sanity(fit1)

# look at the  params
tidy(fit1, "ran_pars")
summary(fit1)

# looking for approximately normal along the line
lng_utm$resid <- residuals(fit1, type = "mle-mvn")
qqnorm(lng_utm$resid)
abline(0,1)
# kinda looks less than great. log length + gaussian still not great
# log normal looks worse


# spatial residuals
# here we're revuing the data to make sure we're not seeing patterns of clustering of
# positive and negative residuals together
ggplot() +
  geom_point(data = lng_utm, aes(X, Y, color = resid)) +
  scale_color_gradient2() +
  facet_wrap(~sample_year)
```

```{r eval=FALSE, include=FALSE}
#' est: fixed + spatial + spatiotemporal (use this for maps and indices)
#' est_non_rf - fixed effects only, use this to understand the covariate-driven signal (in our case here this would
#' be the temporal effects)
#' omega_s - spatial random effects use this to see persistent spatial deviations 
#' epsilon_st - spatiotemporal random effects, will tell you/show year-specific anomalies
preds <- predict(fit1)
head(preds)

pred.grid <- predict(fit1, newdata = bs.grid.year, return_tmb_object = T)

pred.grid2 <- predict(fit1, newdata = bs.grid.year.doy, return_tmb_object = T)
get_index(pred.grid2) %>%
  geom_point(aes(sample_year, est)) +
  geom_errorbar(aes(sample_year, ymin = lwr, ymax = upr))

visreg(fit1)
```


# 3.1 Shipboard lng sdm
```{r}
head(lngs)

lngs_corr_utm <- lngs %>%
  # for now we'll remove the na lengths but ideally we'll fix this on the db side so it won't be a prblem
  filter(!(is.na(corrected_length))) %>%
  mutate(date = as.Date(haul_date),
         doy = yday(date)) %>%
  
  arrange(station_id) %>%
  dplyr::select(c(sample_year, station_id, doy, eq_latitude, eq_longitude, corrected_length)) %>%
  add_utm_columns(., ll_names = c("eq_longitude", "eq_latitude"),
                  ll_crs = 4326, utm_names = c("X", "Y"), 
                  units = "km") # uses crs 32603
  
# create mesh
mesh.lng <- make_mesh(lngs_corr_utm, c("X", "Y"), n_knots = 50)
plot(mesh.lng)
```

Plot sample sizes
```{r}
lngs$station_name <- paste(plyr::round_any(lngs$eq_latitude,0.25),plyr::round_any(lngs$eq_longitude,0.25),sep="")


lngs %>%
  group_by(station_name) %>%
  dplyr::mutate(sample.size = n(),
                   ) %>%
  ggplot() +
  # note about longitude need to add 360 to deal with the funkiness at 180 intl date line
  # ** in wgs, if in other crs i dunno
  geom_hex(aes(eq_longitude + 360, eq_latitude, fill = sample.size, group = log(sample.size))) +
  geom_sf(data = bs_cropped, fill = "gray", color = "black", inherit.aes = F) +
  geom_contour(data = bathline, aes(x = x + 360, y = y, z = z),
               breaks = c(-20, -50, -200), linewidth = 0.3, colour = "black") +
  labs(fill = "Number of samples") +
  viridis::scale_fill_viridis(option = "inferno") +
  #coord_sf(crs="EPSG:3476",default_crs=sf::st_crs(4326),xlim=c(-175,-155),ylim=c(54,65)) +
  theme_minimal()
```


fit measured lengths
```{r}
fit.lng <- sdmTMB(
  data = lngs_corr_utm,
  # model formula
  log(corrected_length) ~ 0 + s(doy) + as.factor(sample_year),
  # add mesh
  mesh = mesh.lng, 
  # we want to fit a spatiotemporal model with iid so we need to say which is the time category
  time = "sample_year", 
  # family important.... start with gaussian, then lognormal 
  family = gaussian(link = "identity"),
  #family = lognormal(link = "log"), 
  spatial = "on",
  # each year independentally estimated
  spatiotemporal = "IID",
  # we want ansiotropy because we want to assume that the decomposition of Matern distance isn't
  # necessarily the same EW as NS. Esp cause in the BS we have the shelf patterns that can drive 
  # species distributions
  anisotropy = TRUE
)
```


```{r}
sanity(fit.lng)

# look at the  params
tidy(fit.lng, "ran_pars")
summary(fit.lng)

# looking for approximately normal along the line
lngs_corr_utm$resid <- residuals(fit.lng, type = "mle-mvn")
qqnorm(lngs_corr_utm$resid)
abline(0,1)
# nto my favourite line there's a decent variability along it but none of the sanity checks seem to cause problems

# spatial residuals
# here we're revuing the data to make sure we're not seeing patterns of clustering of
# positive and negative residuals together
ggplot() +
  geom_point(data = lngs_corr_utm, aes(X, Y, color = resid)) +
  scale_color_gradient2() +
  facet_wrap(~sample_year)

```

create prediction
```{r}
# create prediction grid
## create sf object
lng_corr_sf <- lngs_corr_utm %>%
  ungroup() %>%
  dplyr::select(c(X, Y, eq_latitude, eq_longitude)) %>%
  mutate(X = X*1000, Y = Y*1000) %>% # convert to meters for crs 32602
  st_as_sf(coords = c("X", "Y"), crs = 32602)

## create convex hull
convex <- concaveman(lng_corr_sf, concavity = 3)

## add buffer to edges
convex_buff <- st_buffer(convex, dist = 15e3) # dist in meters (we think)
plot(convex_buff) # looks about right

## make grid
# make grid, created a cell size of 10,000 x 10,000 m (10 x 10 km) with poitns in center
# need to check units on this... 
grid <- st_make_grid(convex_buff, cellsize = 10e3, what = "centers") %>%
  st_intersection(convex_buff)

## change back to df
bs.grid <- do.call(rbind, st_geometry(grid)) %>%
  as_tibble() %>% setNames(c("X", "Y")) %>%
  # convert df m back to km
  mutate(X = X/1e3, Y = Y/1e3)
plot(bs.grid)

## replicate df across years and include any other variables we inlcude in our fit of the model
## i.e., depth tmeperature etc
bs.grid.year <- replicate_df(bs.grid, "sample_year", unique(lngs_corr_utm$sample_year))

bs.grid.year.doy <- replicate_df(bs.grid.year, "doy", 249)
```

predict back to observations
```{r}
# takes awhile to run
#visreg(fit.lng, "sample_year")

#pred.grid <- predict(fit.lng, newdata = bs.grid.year.doy, return_tmb_object = T)

median.doy <- lngs_corr_utm %>%
  group_by(station_id) %>%
  summarise(mean.doy = mean(doy)) %>%
  summarise(med = median(mean.doy),
            mean = mean(mean.doy)) # median doy is 249 mean is 250, aug 28th for non leap year days
# extract visreg outputs, here' we're looking at interannual changes

# vr <- visreg(fit.lng, xvar = "sample_year", cond=list(doy = 249),
#        plot = FALSE, scale = "response")


pred.lng <- predict(fit.lng, newdata = bs.grid.year.doy)

# unsure why this has to happen by the Y (latitude) is off by 1000s
pred.lng.sf <- pred.lng %>%
  mutate(Y = Y,
         X = X) %>%
  st_as_sf(coords = c("X", "Y"), crs = 32602) %>%
  st_transform(., crs = 4326) 

pred.lng.wgs <- pred.lng.sf %>%
  st_shift_longitude() %>%
  dplyr::mutate(lat = sf::st_coordinates(.)[,2],
                lon = sf::st_coordinates(.)[,1])
  

# using the output from predict
ggplot(pred.lng, aes(X, Y, fill = est)) +
  scale_fill_viridis_c() +
  geom_tile() +
  facet_wrap(~sample_year, ncol = 4)


# what might be helpful here too is the average spatial field 
ggplot() +
  geom_sf(data = bs_cropped, fill = "gray", color = "black", inherit.aes = F) +
  # geom_contour(data = bathline, aes(x = x, y = y, z = z),
  #              breaks = c(-20, -50, -200), linewidth = 0.3, colour = "black") +
  geom_sf(mapping = aes(fill = omega_s, color = omega_s),
          data = pred.lng.wgs, pch = 22) +
  # geom_sf(mapping = aes(fill = omega_s), data = pred.lng.sf) +
  scale_fill_viridis_c() +
  scale_color_viridis_c()
```



# 3.2 2024 SDM

```{r}
lngs24 <- lngs %>%
  # for now we'll remove the na lengths but ideally we'll fix this on the db side so it won't be a prblem
  filter(!(is.na(corrected_length))) %>%
  filter(sample_year == "2024",
         region == "SEBS") %>%
  mutate(date = as.Date(haul_date),
         doy = yday(date),
         scaledoy = scale(doy, center = TRUE, scale = TRUE)) %>%
  
  arrange(station_id) %>%
  dplyr::select(c(sample_year, station_id, doy, scaledoy, eq_latitude, eq_longitude, corrected_length)) %>%
  add_utm_columns(., ll_names = c("eq_longitude", "eq_latitude"),
                  ll_crs = 4326, utm_names = c("X", "Y") , 
                  utm_crs = 26902, # trying a NAD83 error
                  units = "km") # uses crs 32603
  
# create mesh
mesh24 <- make_mesh(lngs24, c("X", "Y"), n_knots = 15)
plot(mesh24)
```

```{r}
lngs24$station_name <- paste(plyr::round_any(lngs24$eq_latitude,0.25),plyr::round_any(lngs24$eq_longitude,0.25),sep="")

station.pts <- lngs24 %>%
  group_by(station_name) %>%
  dplyr::mutate(sample.size = n(),
                   )
  
ggplot() +
  # note about longitude need to add 360 to deal with the funkiness at 180 intl date line
  # ** in wgs, if in other crs i dunno
  geom_hex(aes(eq_longitude + 360, eq_latitude, fill = sample.size, group = log(sample.size)),
           data = station.pts) +
  geom_sf(data = bs_cropped, fill = "gray", color = "black", inherit.aes = F) +
  geom_contour(data = bathline, aes(x = x + 360, y = y, z = z),
               breaks = c(-20, -50, -200), linewidth = 0.3, colour = "black") +
  labs(fill = "Log Number of samples") +
  viridis::scale_fill_viridis(option = "inferno") +
  #coord_sf(crs="EPSG:3476",default_crs=sf::st_crs(4326),xlim=c(-175,-155),ylim=c(54,65)) +
  theme_minimal()
```

## create prediction grid
```{r}
# create prediction grid
## create sf object
lngs24.sf <- lngs24 %>%
  ungroup() %>%
  dplyr::select(c(X, Y, eq_latitude, eq_longitude)) %>%
  mutate(X = X*1000, Y = Y*1000) %>% # convert to meters for crs 32602
  st_as_sf(coords = c("X", "Y"), 
           crs = 26902 )
           #crs = 32602)

## create convex hull
convex <- concaveman(lngs24.sf, concavity = 3)

## add buffer to edges
convex_buff <- st_buffer(convex, dist = 15e3) # dist in meters (we think)
plot(convex_buff) # looks about right

## make grid
# make grid, created a cell size of 10,000 x 10,000 m (10 x 10 km) with poitns in center
# need to check units on this... 
grid <- st_make_grid(convex_buff, cellsize = 10e3, what = "centers") %>%
  st_intersection(convex_buff)

## change back to df
bs.grid <- do.call(rbind, st_geometry(grid)) %>%
  as_tibble() %>% setNames(c("X", "Y")) %>%
  # convert df m back to km
  mutate(X = X/1000, Y = Y/1000)
plot(bs.grid)

## replicate df across years and include any other variables we inlcude in our fit of the model
## i.e., depth tmeperature etc
bs.grid.doy <- replicate_df(bs.grid, "scaledoy", 0) # between 235 and 236 is ~0

# lngs24 %>%
#   summarise(med = median(doy),
#             mean = mean(doy)) 
```

## fit 2024 model
because this is not a temporal model, basically what we want to do is use samples to create a spatial 
extrapolation of length from 2024 while accounting for doy
```{r}
fit.lng24 <- sdmTMB(
  data = lngs24,
  # model formula
  log(corrected_length) ~ s(scaledoy),
  # add mesh
  mesh = mesh24, # knot 15
  # we want to fit a spatiotemporal model with iid so we need to say which is the time category
  # time = "sample_year", 
  # family important.... start with gaussian, then lognormal 
  family = gaussian(link = "identity"),
  #family = lognormal(link = "log"), 
  spatial = "on",
  # each year independentally estimated
  #spatiotemporal = "IID",
  # we want ansiotropy because we want to assume that the decomposition of Matern distance isn't
  # necessarily the same EW as NS. Esp cause in the BS we have the shelf patterns that can drive 
  # species distributions
  anisotropy = FALSE
)

run_extra_optimization(fit.lng24, newton_loops = 3)

sanity(fit.lng24)
summary(fit.lng24)
tidy(fit.lng24)
```


```{r}
# note we used a gaussian link identity for the length of this but log transformed it
pred24 <- predict(fit.lng24, newdata = bs.grid.doy) %>%
  mutate(Y = Y * 1000, 
         X = X * 1000) # convert to meteres because sf assumes meters

ggplot(pred24, aes(X, Y, fill = exp(est))) +
  scale_fill_viridis_c() +
  geom_tile() 


bs_cropped_shift <- bs_cropped %>%
  st_transform(crs = 4326) %>%
  st_shift_longitude() #%>%
  # st_crop(ymin = 54, ymax = 68, xmin = -180, xmax = -150)
  

pred24.sf <- st_as_sf(pred24, coords = c("X", "Y"), 
                      crs = 32602
                      #crs = 26902
                      ) %>%
  st_transform(crs = st_crs(bs_cropped_shift)) %>%
  st_shift_longitude()

st.pts.sf <- station.pts %>%
  st_as_sf(coords = c("eq_longitude", "eq_latitude"), crs = 4326) %>%
  st_transform(crs = st_crs(bs_cropped_shift)) %>%
  st_shift_longitude()


ggplot() +
  geom_sf(aes(fill = exp(est), color = exp(est)), pch = 22, data = pred24.sf) +
  scale_fill_viridis_c() + scale_color_viridis_c() +
  geom_sf(data = bs_cropped_shift, fill = "gray", color = "black", inherit.aes = F) +
  geom_contour(data = bathline.xyz, aes(x = V1 +360, y = V2, z = V3, label),
               breaks = c(-20, -50, -70, -150), linewidth = 0.3, colour = "black") +
  # geom_sf(aes(fill = sample.size, group = log(sample.size)), pch = 22,
  #          data = st.pts.sf) +
  theme_bw() +
  # gets rid of the span between the edge of the plot and axes
  coord_sf(xlim = c(185, 202), ylim = c(54, 61), expand = FALSE, crs = 4326) +
  labs(x = "Longitude", y = "Latitude", color = "Predicted length (mm)", fill = "Predicted length (mm)")


```

